\section{Scientific model}
\subsection{Speed distribution}

The well-known \textbf{Maxwell-Boltzmann distribution} describes the molecular 
speed for a monoatomic classical ideal gas in thermodynamic equilibrium.

$$
f(v)\dif v = \left( \frac{\beta m}{2 \pi} \right)^{3/2} 4 \pi v^2 e^{-\beta 
\frac{mv^2}{2}} \dif v
$$
where $\beta$ is given by
\begin{math}
\beta \equiv \frac{1}{k_BT}
\end{math}
, $T$ is the temperature, $k_B$ is the Boltzmann constant and $m$ is the 
molecular mass of the gas.

In this section, we derive a similar expression for a two-dimensional gas.
Consider a gas consisting of a single molecule of mass $m$ in a container in 
contact with a heat bath of temperature $T$. The probability of the particle 
being in a state with energy $E$ is given by the Boltzmann distribution.
$$
P(E) \dif E \propto e^{-\beta E}\dif E
$$
Since the energy of particle is only a function of the velocity $\vec{v}$, $E = 
\frac{mv^2}{2}$, we can deduce
$$
P(\vec v)\dif{\vec v} \propto e^{- \beta \frac{m |\vec v|^2}{2}}\dif\vec v
$$
But we're interested in the speed distribution, not the velocity distribution.  
For this, we use a substitution to polar coordinates
$$
\dif \vec{v} = \dif v_x\dif v_y = v\dif v\dif \theta
$$
where $v=|\vec{v}|$ is the speed. So
$$
f(v)\dif v\dif \theta \propto ve^{- \beta \frac{mv^2}{2}} \dif v\dif \theta
$$
Since the distribution is independent of the parameter $\theta$, we can 
integrate over it and remove it from the equation.
$$
f(v)\dif v\propto ve^{- \beta \frac{mv^2}{2}} \dif v
$$
After normalisation we find the \textbf{Rayleigh distribution}
$$
f(v)\dif v=\beta m ve^{-\beta \frac{mv^2}{2}} \dif v
$$

\subsection{Collisions}

The most basic short-range interaction that can be modeled in a many particle 
system is the elastic collision between hard spheres or disks. It is important 
that this interaction is not only addressed correctly, but also efficiently.  
Therefore, we want to avoid calculating computationally expensive trigonometric 
functions like sine and cosine and try to keep square roots to a minimum.

Collision simulations function in basically two ways, where the collisions are 
detected before they happen, \emph{a priori}, or after, \emph{a posteriori}.  

In the a posteriori case, we advance the simulation by a small step each 
iteration and check if collisions have happened. If they did, we roll the 
simulation back a short while to before the collision happened and restore the 
situation correctly. In the a priori case, our collision detection algorithm 
needs to interpolate the positions of various bodies to predict whether a 
collision will happen or not.

A priori algorithms are more stable and correcter, because the intersections of 
bodies that need to be fixed in the a posteriori case are unphysical. However, 
determining ahead of time when two bodies will collide is very difficult and it 
requires the collision detection algorithm to know about the physics of moving 
bodies.

Since we detect collisions a posteriori, we need to restore the situation to 
before the collision. Thus, we need to rewind each pair of intersecting spheres 
to before the collision. Assuming there is no external forcefield, the 
particles are in a state of uniform rectilinear motion. The equation we need to 
solve for the time $t$ is
$$
|\vec{r_1}(t) - \vec{r_2}(t)| = R_1 + R_2
$$
with $\vec{r_i}$ the position vectors of the bodies and $R_i$ their radii.  
Since their motion is uniform, this becomes
$$
|\vec{r_1}{t_0} + \vec{v_1}\Delta t - \vec{r_2}{t_0} - \vec{v_2}\Delta t| = | 
\Delta\vec{r} + \Delta\vec{v} \Delta t | = R_1 + R_2
$$
with $\Delta\vec{r}$ and $\Delta\vec{v}$ the particles' relative position and 
motion and $\Delta t$ the sought after time difference. Squaring both sides, we 
find
$$
(\Delta\vec{r})^2 + 2\Delta\vec{v}\Delta\vec{r}\Delta t + 
(\Delta\vec{v})^2(\Delta t)^2 = (R_1 + R_2)^2
$$
The solutions of this quadratic equation are
$$
\Delta t = \frac{-2\Delta\vec{v}\Delta\vec{r} \pm 
\sqrt{4(\Delta\vec{v}\Delta\vec{r})^2 - 4(\Delta\vec{v})^2((\Delta\vec{r})^2 - 
(R_1 + R_2)^2)}}{2(\Delta\vec{v})^2}
$$
As $\Delta\vec{v}\Delta\vec{r}$ is typically a negative number and $R_1 + R_2$ 
is always greater than $|\Delta\vec{r}|$, we take the positive squareroot and 
simplify
$$
\Delta t=\frac{-\Delta\vec{v}\Delta\vec{r} + 
\sqrt{(\Delta\vec{v}\Delta\vec{r})^2 + 
(\Delta\vec{v})^2((R_1+R_2)^2-(\Delta\vec{r})^2)}}{(\Delta\vec{v})^2}
$$
With this value, we can put the particles back in their original position 
before the collision. Now, we need to handle the collision itself.

\subsubsection{One-dimensional elastic collision}

Consider two particles denoted by subscripts 1 and 2.  Let $m_i$ be the masses, 
$v_c$ be the the velocity of the center of mass (COM) frame, $v_i$ the 
velocities before the collision, $v'_i$ the velocities before collision in the 
COM frame, $w_i$ the velocities after the collision and $w'_i$ the velocities 
after collision in the COM frame.  We assume the speed of both bodies is 
non-relativistic. 

To considerably simplify the equations, we are going to transform to the 
intertial frame of the center of mass (COM). The COM velocity is
$$
v_c = \frac{m_1 v_1 + m_2 v_2}{m_1 + m_2}
$$

The resulting velocities in the new frame are
$$
v'_1 = v_1 - v_c = \frac{m_2}{m_1+m_2}(v_1-v_2)
$$
and
$$
v'_2 = v_2 - v_c = \frac{m_1}{m_1+m_2}(v_2-v_1)
$$

In this frame, the total momentum must be zero before and after the collision.  
The kinetic energy must also be conserved.
Thus
$$
p_1 + p_2 = q_1 + q_2 = 0
$$
and
$$
\frac{p_1^2}{2m_1} + \frac{p_2^2}{2m_2} = \frac{q_1^2}{2m_1} + 
\frac{q_2^2}{2m_2}
$$
where $p_i$ and $q_i$ are the momenta respectively before and after the 
collision.

From this follows that
$$
q_2 = -q_1
$$
and
$$
p_2 = -p_1
$$
which we can fill in in the energy equation
$$
p_1^2 = q_1^2
$$
and
$$
p_2^2 = q_2^2
$$
The only possible solutions are $q_i = p_i$, that is no collision happened at 
all, or $q_i = -p_i$. This means that the velocities of both bodies are 
flipped.

$$
w'_i = -v'_i
$$

Thus we can derive the velocities in the original frame of reference
$$
w_1 = w'_1 + v_c = \frac{v_1(m_1-m_2) + 2m_2u_2}{m_1+m_2}
$$

$$
w_2 = w'_2 + v_c = \frac{v_2(m_2-m_1) + 2m_1u_1}{m_1+m_2}
$$

\subsubsection{Two- and higher-dimensional elastic collisions}
Because the colliding bodies are spheres, the only forces they can exert on 
each other at the contact point are according to the normal vector of each 
sphere at the contact point. The tangential component(s) of the momentum of 
each body do not change while the normal component is transformed according to 
the equations of an elastic collision. To get the normal vector, we simply 
normalise the displacement vector between the bodies
$$
\vec{d} = \frac{\Delta\vec{r}}{|\Delta\vec{r}|}
$$

\subsection{Spatial partitioning}

Most many-particle simulations incorporate some sort of interaction between the 
particles dependent on their mutual distance. Of considerable interest are 
short-range interactions, for example when the particles represent atoms. Since 
a particle could hypothetically interact with every other particle, each pair 
needs to be tested separately. Thus $n(n-1)/2$ interactions will be considered, 
with $n$ the number of particles. Most of these are useless because the 
particles are too far apart. This $O(n^2)$ complexity is undesirable for 
performance reasons. An often used technique is called \textbf{spatial 
partitioning}.

By splitting the world into smaller partitions,\emph{boxes}, one only has to 
check for interactions between particles in nearby partitions. If we keep the 
particles per box a constant ($x$), the amount of boxes becomes $n/x$. The 
complexity of this problem reduces in this way to $O(x^2 * n/x) = O(n)$, linear 
in the number of particles.
